---
title: "stabilitest: Robustness and Fragility Analysis of Statistical Test Conclusions in R"
date: "2026-08-07"
abstract: >
  A p-value quantifies how surprising the observed data would be under the null hypothesis, but says nothing about how easily the resulting conclusion could be overturned. Regulatory guidance (ICH E9(R1)) asks sponsors to demonstrate that trial conclusions are robust to data limitations and analytic assumptions, without prescribing quantitative machinery. The stabilitest package addresses this gap by combining three complementary sensitivity views of a single pre-specified analysis: jackknife leave-one-out influence, greedy worst-case observation removal in the spirit of the maximum influence perturbation of Broderick, Giordano and Meager, and bootstrap reproducibility probability. Component metrics and a composite 0-100 score are returned for two-sample location and proportion tests, linear model and ANCOVA terms, GLM and Cox terms, and TOST equivalence and non-inferiority endpoints. The package's distinguishing feature is its treatment of categorical interpretation as a claim requiring evidence: a Robust or Fragile label is emitted only where an independent, pre-registered calibration study has validated thresholds for that exact analysis configuration, and is otherwise suppressed while numeric output is retained. Two such studies are reported, one positive (a validated two-band mapping for Fisher's exact test) and one negative across three attempts (ANCOVA, where the composite score proves to be a near-monotone re-expression of the p-value and no feasible thresholds exist). The negative result is presented as evidence for the fail-closed design rather than omitted.
draft: false
author:
  - name: Marius Ardelean
    affiliation: Independent researcher
    email: marally@gmail.com
    url: https://github.com/ma-brain/stabilitest
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
header-includes:
  - \providecommand{\pandocbounded}[1]{#1}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 4.2, out.width = "100%"
)
library(stabilitest)
library(ggplot2)

# Heavy evidence is loaded from committed artifacts generated by the scripts in
# tools/ with the released package; see REPRODUCE.md. Live chunks below are
# limited to short worked examples so the article knits well inside the
# journal's 10-minute budget.
art <- function(...) file.path("artifacts", ...)
case_study  <- readRDS(art("case-study", "welch-case-study.rds"))
sim_summary <- readRDS(art("simulation", "simulation-summary.rds"))
timing      <- readRDS(art("timing", "timing.rds"))
test_counts <- readRDS(art("testing", "test-counts.rds"))

fmt <- function(x, digits = 1) formatC(x, format = "f", digits = digits)
cs_m <- case_study$robustness_metrics
```

# Introduction

Statistical reporting in clinical trials typically compresses the evidence into
a binary decision at $p < \alpha$. That compression discards something a
reader needs. Consider two trials that both report $p = 0.048$. In the first,
the conclusion survives essentially any plausible perturbation of the data. In
the second, it rests entirely on one extreme responder: delete that single
patient and significance evaporates. Standard reporting treats these as
identical findings.

This is not a hypothetical concern: sensitivity of conclusions to individual
patients is a recurring theme in the trial-reporting literature
[@senn2007; @pocock2015]. The ICH E9 guideline and its E9(R1) addendum
[@ich1998e9; @ich2019e9r1] make it explicit, asking that inferences be robust
to data limitations and to deviations from the assumptions of the main
estimator, assessed through pre-specified sensitivity analyses. What the
guidance does not supply is quantitative machinery. Three distinct questions jointly characterise the
stability of a test conclusion, and each needs a different tool:

1. *Which individual observations drive the result?* (influence)
2. *What is the smallest set of observations whose removal would overturn the
   conclusion?* (fragility)
3. *Would a replicate sample likely reach the same conclusion?*
   (reproducibility)

\CRANpkg{stabilitest} [@ally2026stabilitest] answers all three for one
pre-specified primary analysis, returning component metrics plus a composite
score. Its contributions are:

- **An integrated framework.** Jackknife influence, adversarial worst-case
  removal, and bootstrap reproducibility are computed together for a common
  catalogue of trial-relevant analyses --- two-sample location and proportion
  tests, linear model and ANCOVA terms, GLM and Cox terms, and TOST
  equivalence and non-inferiority --- through a small, uniform API.
- **A fail-closed calibration registry.** Categorical labels
  (*Robust* / *Fragile*) are treated as claims that require evidence. The
  package emits a label only where an independent, pre-registered calibration
  study has validated thresholds for that exact method, endpoint, conclusion
  type, weight vector, and design profile. Everywhere else the label is `NA`
  while numeric output is retained. Thresholds are never transferred between
  method families by analogy.
- **A calibrated unit, earned.** An independent study for Fisher's exact test
  yielded a validated two-band mapping, reported here with its held-out
  confirmation bounds.
- **A named negative result.** Three pre-registered attempts to calibrate
  ANCOVA all closed without thresholds. The mechanism is informative rather
  than merely disappointing: in clean parametric settings the composite score
  is close to a monotone re-expression of the $p$-value, so no score built from
  the same data can deliver the discrimination the acceptance gates require.
  This is reported as a contribution, not omitted.

# Related R packages

Every package in Table \@ref(tab:related) answers a version of "how much would
have to change for this conclusion to flip?", but they fall into two families.

\CRANpkg{sensemakr} [@cinelli2020], \CRANpkg{konfound} and \CRANpkg{EValue}
ask the question about *unmeasured* threats --- a confounder, a selection
mechanism, or unobserved cases the analyst never had. Their answers are
analytic bounds computed from the reported estimate and its uncertainty; none
of them touches the rows of the analysis dataset. \CRANpkg{fragility},
\CRANpkg{influence.ME}, \CRANpkg{car}, `zaminfluence` (GitHub only, from the
authors of @broderick2023) and \pkg{stabilitest} instead ask about the
*observed* sample: which specific patients, if perturbed or removed, would
change the conclusion?

Within that second family the differences are of scope and rigour rather than
kind. \pkg{fragility} is restricted to flipping binary outcome labels, implementing
the clinical fragility index of @walsh2014. \pkg{car} and \pkg{influence.ME}
offer classical single-deletion diagnostics in the tradition of @cook1977,
which the simulation in the Calibration section shows can overstate stability
by a factor of two to five relative to adversarial removal. `zaminfluence`
approximates the adversarial leave-$k$-out answer efficiently via influence
functions, but for Z-estimators in general rather than the two-sample and
model-term designs of trial reporting.

```{r related}
related <- data.frame(
  Package = c("fragility", "sensemakr", "konfound", "EValue", "boot",
              "influence.ME", "car", "zaminfluence"),
  `Question answered` = c(
    "Minimal outcome-status flips reversing significance",
    "Strength an omitted confounder needs to explain away a coefficient",
    "Bias or share of cases needed to invalidate an inference",
    "Strength unmeasured bias needs to explain away an association",
    "General-purpose bootstrap resampling of any statistic",
    "Which grouping units are influential in an lme4 fit",
    "Outliers, leverage and single-deletion influence on lm/glm",
    "Approximate minimal removal fraction for a Z-estimator"),
  `Endpoint scope` = c(
    "Binary only; studies and meta-analyses",
    "Linear regression coefficients",
    "General regression-style inferences",
    "Observational effect estimates",
    "Any user-supplied statistic",
    "Mixed models; grouping factors",
    "Any lm/glm fit",
    "Z-estimators (econometric applications)"),
  check.names = FALSE
)
knitr::kable(related, caption = "R packages addressing related sensitivity questions. All entries verified against live CRAN pages or repository documentation on 2026-08-07; fragilityindex was checked and found archived on CRAN since 2020-07-08 and is therefore excluded.", booktabs = TRUE)
```

\pkg{stabilitest}'s distinguishing choices relative to all of these are: it
evaluates deletion, jackknife and bootstrap sensitivity together as three
views of one conclusion rather than offering a single diagnostic; its
worst-case removal is an actual adversarial refit search rather than a
residual/leverage heuristic or a linear approximation; and its composite score
carries a categorical verdict only where an independent pre-registered
calibration has validated it for that exact configuration. No other package in
Table \@ref(tab:related) ships a calibration registry or a fail-closed
labelling policy.

# Methods

Given two samples --- or a model term, see the API section --- and a
significance level $\alpha$, four analyses are performed. Three enter the
composite score; the fourth is retained as a deliberate contrast.

**Jackknife (leave-one-out).** Each observation is deleted in turn and the
test repeated, following the classical framework of @shaotu1995. Outputs are
the conclusion stability $S_\mathrm{jack}$ (percentage of deletions preserving
the conclusion), the set of influential observations (deletion flips the
conclusion, or shifts $p$ by more than $\delta = 0.05$), and the leave-one-out
$p$-value range. Because single-deletion influence shrinks as $\sim 1/n$, high
jackknife stability in a large sample is *weak* evidence of robustness. The
component's chief value is identifying specific observations for data-quality
review; the package's printed output says so explicitly.

**Worst-case removal (primary fragility analysis).** A greedy algorithm
deletes, at each step, the observation whose removal moves the $p$-value
furthest toward overturning the conclusion, up to
$k_{\max} = \lfloor 0.30\,n \rfloor$ deletions. The *removal fragility index*
$k_\mathrm{frag}$ is the step at which the conclusion first flips. This
follows the maximum influence perturbation of @broderick2023 (see also
@giordano2026a and @giordano2026b), who showed that in prominent published
analyses removing under 1% of observations can reverse the sign of an
estimated effect. The greedy set is *sufficient* to overturn the conclusion, so
$k_\mathrm{frag}$ is an upper bound on the size of the minimal overturning
subset --- in practice a far tighter bound than any symmetric outlier-trimming
rule provides. Because a flip from $p = 0.047$ to $p = 0.053$ is a knife-edge
event [@potter2020], the package returns and plots the full $p$-value
trajectory rather than the index alone. Note the terminology: the *removal*
fragility index counts deleted observations and so reduces $n$, which
distinguishes it from the Walsh fragility index [@walsh2014], which flips
event status while holding $n$ fixed.

**Extreme-value removal (descriptive).** Cumulative deletion of observations
most distant from the recomputed grand mean --- the familiar "we removed the
outliers and the result held" procedure. It does *not* enter the composite
score. It is retained because the gap between this index and the worst-case
index measures how much protection the outlier narrative actually provides.

**Bootstrap reproducibility.** $B$ resamples drawn with replacement within
groups [@efron1993]; $S_\mathrm{boot}$ is the proportion reaching the original
conclusion. Because resampling preserves the observed effect, this estimates
the *reproducibility probability* [@goodman1992; @shao2002] --- approximately
the power at the observed effect size. It reflects strength of evidence, not
resistance to contamination: a marginal $p$-value yields $S_\mathrm{boot}$
near 50--60% even in perfectly clean data. It is reported with that framing
and weighted lightly.

**Composite score.**
$R = w_1 S_\mathrm{jack} + w_2 F + w_3 S_\mathrm{boot}$, where
$F = 100 \cdot \min\!\left(k_\mathrm{frag} / (k_{\max}+1),\, 1\right)$
rescales the fragility index to the full 0--100 range. Default weights are
$w = (0.4, 0.4, 0.2)$: influence and fragility carry equal weight, the
reproducibility term is down-weighted because it largely restates the
$p$-value. The weights are a heuristic communication device, not a derived
quantity. They are an explicit argument and should be fixed in the statistical
analysis plan. Two results with the same score can be stable in different
ways, so the components --- especially the fragility index and the identity of
the observations behind it --- are the substance of reporting.

# Package design and API

## Dispatcher and engines

Two-sample analyses go through a single dispatcher; model-based and
equivalence analyses have their own entry points sharing a common
case-deletion engine.

```{r api-demo, echo=TRUE, eval=FALSE}
# Two-sample: t.test, paired.t.test, wilcoxon, wilcoxon.paired,
#             brunner_munzel, fisher, chisq, prop
robustness_analysis(group1, group2, test_type = "t.test")

# Model terms: single coefficient or a multi-df factor tested jointly
robustness_lm(change ~ arm + baseline, data = d, term = "arm")
robustness_glm(y ~ arm + age, data = d, term = "arm", family = binomial())
robustness_surv(Surv(time, event) ~ arm, data = d, term = "arm")

# Equivalence / non-inferiority: mean, risk difference, odds ratio
robustness_tost(g1, g2, type = "equivalence", endpoint = "mean",
                delta_L = -20, delta_U = 20)
```

For model engines all components operate on *rows* (subjects) of the analysis
dataset, with the conclusion defined by the $p$-value of a pre-specified term
--- a single coefficient, or a multi-df factor term via a joint test
(`drop1` $F$ for `lm`, likelihood-ratio for `glm` and `coxph`). Deleting whole
subjects, with their covariates and follow-up, preserves the adjustment
structure and handles censoring naturally. Fits that fail after deletion
(a factor level vanishing, say) are skipped and reported rather than silently
dropped.

Arguments are deliberately uniform across engines: `alpha`, `n_boot`,
`max_removal_pct`, `weights`, and `seed` mean the same thing everywhere, and
`seed` defaults to `123` so that a bare call is reproducible.

## Return objects and methods

Analyses return a classed list (`robustness_analysis`, `robustness_model`, or
`robustness_tost`) containing the original test result, a
`robustness_metrics` tibble of 17 component metrics, the per-step trajectories
of each removal analysis, the bootstrap draws, the resolved calibration
decision, and the analysis profile used to make that decision. `print()`
methods render a structured summary; `plot()` returns the removal trajectories
and the bootstrap $p$-value distribution.

## The calibration registry

The mechanism that makes label suppression systematic rather than ad hoc is a
registry shipped as `inst/extdata/calibration-registry.csv`, keyed by the
*resolved* method rather than by the user-facing argument. A call with
`test_type = "t.test"` resolves to `welch_unpaired` or `paired_t`; one with
`test_type = "fisher"` resolves to `fisher_exact`. Each row carries a status,
cutoffs where applicable, a version string, and a provenance pointer to the
study that produced it.

Label resolution is fail-closed. A categorical label is emitted only when the
resolved method has a `validated_method_specific` row **and** the observed
conclusion type matches **and** the weight vector matches the calibrated one
**and** the design profile falls inside the study's eligibility bounds. Any
mismatch yields `NA` --- with a printed reason --- while the numeric score and
all components are retained. There is no fallback to a "nearest" calibration:
the Welch 55/70 thresholds are a Welch result and are never applied to ANCOVA.

# Calibration policy and evidence

## The Welch unit

The framework's original calibration targets the significant `welch_unpaired`
configuration under default weights, with three bands: score $> 70$ is
*Robust*, $(55, 70]$ is *Moderately Robust*, $\leq 55$ is *Fragile*.

Table \@ref(tab:simtable) reports the supporting simulation, regenerated for
this article with the released package (12 scenarios $\times$ 500
replications, $B = 200$; see REPRODUCE.md). Conditional on significance, the
separation is stark: null-data findings that reached significance by chance
score around 50--55 and collapse after removing one or two observations, while
genuine large effects ($d = 0.8$, $n = 50$) score 75--78 and withstand removal
of 12--14 observations.

```{r simtable}
st <- sim_summary[, c("d", "n_per_group", "n_outliers", "rejection_rate",
                       "score_sig", "k_wc_med_sig", "k_ex_med_sig", "s_boot_sig")]
st$score_sig  <- round(st$score_sig, 1)
st$s_boot_sig <- round(st$s_boot_sig, 1)
names(st) <- c("$d$", "$n$", "Cont.", "Rej. rate", "Score",
                "$k_{\\mathrm{wc}}$", "$k_{\\mathrm{ex}}$", "Boot. \\%")
knitr::kable(st, digits = 3, row.names = FALSE, booktabs = TRUE, escape = FALSE,
  caption = "Simulation evidence for the Welch calibration, conditional on statistical significance. 500 replications per scenario, B = 200, regenerated with stabilitest 0.6.0. Columns: effect size, per-group n, number of contaminating observations injected at +4 SD into the treated group, rejection rate, mean composite score among significant results, median worst-case and extreme-value fragility indices, and mean bootstrap reproducibility.")
```

```{r simclaims}
n0 <- sim_summary[sim_summary$d == 0, ]
rej_clean <- n0$rejection_rate[n0$n_outliers == 0]
rej_cont  <- n0$rejection_rate[n0$n_outliers == 2]
c50 <- sim_summary[sim_summary$d == 0.8 & sim_summary$n_per_group == 50 &
                     sim_summary$n_outliers == 0, ]
mc_se <- sqrt(0.05 * 0.95 / 500)
```

Two features of Table \@ref(tab:simtable) deserve emphasis. First, under the
null the rejection rate is at or near nominal in clean data
(`r paste(sprintf("%.3f", rej_clean), collapse = " and ")` across the two
sample sizes) and is higher under contamination
(`r paste(sprintf("%.3f", rej_cont), collapse = " and ")`). The Monte Carlo
standard error at 500 replications is about `r fmt(mc_se, 3)`, so individual
cells should not be over-read --- the Welch test's type I error is of course
nominal by construction. What matters for calibration is not the rejection
rate but what the components say about the findings that *did* reach
significance: their median worst-case fragility is one or two observations
throughout the null rows, which is exactly the signature the bands are meant
to catch.

Second, and more consequentially, the median extreme-value index is
systematically larger than the worst-case index under a genuine effect ---
`r c50$k_ex_med_sig` versus `r c50$k_wc_med_sig` at $d = 0.8$, $n = 50$.
Judging stability by removing apparent outliers makes a result look roughly
twice as sturdy as an adversarial reading shows it to be. The most damaging
observations are usually not the obvious ones.

Figure \@ref(fig:simfig) shows the underlying score distributions.

```{r simfig, fig.cap="Composite score distributions for significant results under the null (d = 0, left) and a large effect (d = 0.8, right), n = 50 per group, uncontaminated. Dashed lines mark the calibrated Welch bands at 55 and 70. The separation is what licenses the bands; the overlap is why the bands are three-way rather than binary.", fig.height=3.4}
s_null  <- readRDS(art("simulation", "scenario-03.rds"))
s_clear <- readRDS(art("simulation", "scenario-11.rds"))
dd <- rbind(
  data.frame(score = s_null$score[s_null$significant],   truth = "Null (d = 0)"),
  data.frame(score = s_clear$score[s_clear$significant], truth = "Large effect (d = 0.8)")
)
dd$truth <- factor(dd$truth, levels = c("Null (d = 0)", "Large effect (d = 0.8)"))
ggplot(dd, aes(score)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  geom_vline(xintercept = c(55, 70), linetype = "dashed") +
  facet_wrap(~truth, scales = "free_y") +
  labs(x = "Composite robustness score", y = "Count") +
  theme_minimal()
```

## The Fisher unit: a calibration earned

An independent study for Fisher's exact test on two-arm binary outcomes
delivered the package's second active calibration. Its design was frozen
before execution: truth strata defined by *enumerated exact power* rather than
by observed effect size (null exact; clear at exact power 0.95; a borderline
stratum at 0.60 retained as diagnostic only and excluded from threshold
fitting), and a sealed score-only pilot gating production.

The study also settled the weight question empirically. On binary data the
jackknife stability component is constant at 100 in every cell --- a single
deletion cannot move a Fisher $p$-value across $\alpha$ at these sample sizes
--- so it carries zero information. The frozen design therefore uses
*jackknife-light* weights $(0, 0.5, 0.5)$, which maximise score granularity
without changing discrimination.

Training selected the cutoff $L = 58$ with false reassurance 0.0493 (Wilson
upper 0.0607) and identification of clear effects 0.7608 (Wilson lower 0.740).
Held-out confirmation, opened only once, gave 0.0367 and 0.7333; under the
conservative-of-Wilson-and-cluster-bootstrap rule the reported bounds are a
false-reassurance upper of 0.0617 and an identification lower of 0.6767, both
inside the frozen gates. The registry row is `fisher-2026-1`, provenance
`study:binary_proportion@cc3344931614`.

The resulting mapping is two-band --- score $\leq 58$ *Fragile*, $> 58$
*Not fragile* --- with **no Robust tier**, because a single binary comparison
does not carry enough information to support a third band at the required
error rates. It applies only within the study's eligibility profile: per-arm
$n$ in $[25, 200]$, allocation ratio in $[0.8, 1.25]$, control-arm event rate
in $[0.08, 0.55]$ with at least three events and three non-events. Outside
that profile, or under default weights, the label is suppressed.

## ANCOVA: a named negative result

Covariate-adjusted analysis is the natural next calibration target --- most
confirmatory trials adjust for baseline. Three pre-registered attempts were
made. All three closed without a usable label, and the reason is structural.

**v1** (three-band, default weights) closed fail-closed at Gate B with reason
`no_feasible_thresholds`; held-out data was never opened. Follow-up analysis
of the training data explained why: the composite score is very nearly a
monotone re-expression of the $p$-value (the bootstrap component correlates
with $-\log_{10} p$ at Spearman $\approx 0.99$), and in the canonical Gaussian
ANCOVA design $|t|$ is sufficient. No resampling-based score can beat
$p$-banding when it is a function of the same data. The acceptance gates
implicitly demand an AUC of about 0.94; truth classes defined at 90% power
deliver about 0.89. The infeasibility is information-theoretic, not a
weighting problem.

**v2** (two-band, jackknife-light weights, mirroring the design that later
succeeded for Fisher) was executed before that analysis landed, and failed
exactly as it predicted. Its sealed pilot returned a GO on location metrics
($\Delta = 24.4$, overlap 0.034, AUC 0.892), but training found no feasible
cutoff: the best identification rate at the false-reassurance-safe threshold
was 0.554 against a gate of 0.70. The analytic projection had predicted 0.542.
This makes the episode a *prediction followed by confirmation*: the pilot
gate's location metrics are a false-GO indicator, and the
projected-identification-at-safe-threshold metric that replaced them is now
validated to about 0.01 accuracy.

**v3 Track E** re-aimed at a target where resampling should genuinely add
information beyond $p$: detecting analyses whose assumptions have been
violated --- the one job a $p$-value cannot do. The pre-registered gate
required $\Delta\mathrm{AUC} = \mathrm{AUC}_\mathrm{score} -
\mathrm{AUC}_p \geq 0.10$ with a bootstrap CI excluding zero. Quotas were met
(2,100 replications completed, 0 failed). The result was **not confirmed**:
pooled $\Delta\mathrm{AUC} = 0.0053$, cluster-bootstrap 95% CI
$[-0.111, 0.129]$.

All three registry rows remain `uncalibrated` with auditable provenance, and
ANCOVA labels stay suppressed. Welch 55/70 is a Welch comparator and is never
an ANCOVA fallback. We report this in full because the alternative --- quietly
loosening acceptance criteria until something validates --- is precisely the
failure mode the fail-closed design exists to prevent. Two rigorously
documented negative outcomes plus a confirmed methodological prediction are
evidence that the package's positive claims were earned rather than assumed.

# Examples

## Welch case study

The package ships a synthetic phase II analgesic trial: 55 patients, change
from baseline in pain score at week 12. This is the documented Welch
configuration, so a label applies. Results below are loaded from a committed
artifact computed with package defaults.

```{r case-code, echo=TRUE, eval=FALSE}
res <- robustness_analysis(pain_treatment, pain_placebo, test_type = "t.test",
                           n_boot = 2000, seed = 123, interpret = TRUE)
```

The Welch test gives a mean difference of `r fmt(case_study$original_mean_diff, 2)`
points, $p = `r fmt(case_study$original_p, 4)`$. The robustness analysis
returns a score of **`r fmt(cs_m$overall_robustness)`/100 (Robust)**, built
from: jackknife stability `r fmt(cs_m$jackknife_conclusion_stability)`% with
`r cs_m$jackknife_n_influential` influential observations flagged; a
worst-case fragility index of $k = `r cs_m$worstcase_fragility_k`$
(`r fmt(cs_m$worstcase_fragility_pct)`% of the sample), with $p$ rising to
`r fmt(cs_m$p_at_fragility, 4)` after the sixth deletion; and bootstrap
reproducibility `r fmt(cs_m$bootstrap_reproducibility)`%.

Figure \@ref(fig:casefig) shows the two diagnostic displays produced by
`plot()`. The upper panel makes the contrast of the Methods section concrete:
the worst-case trajectory crosses $\alpha$ well before the extreme-value
trajectory does.

```{r casefig, fig.cap="Diagnostic plots for the Welch case study. Top: p-value trajectories under worst-case (adversarial) and extreme-value (outlier-ranked) removal, with the dashed line at alpha = 0.05. Bottom: bootstrap p-value distribution, red line at the observed p-value.", fig.height=5.5}
invisible(plot(case_study))
```

For a reader, the actionable output is not the score but the six named
patients --- led by one extreme responder --- whose charts warrant source-data
review, and the observation that $k = 6$ matches the validated signature of a
genuine effect rather than the one-or-two signature of a chance finding.

## Other engines

Short live examples across the remaining engines. An ANCOVA term on the
packaged illustration dataset:

```{r ancova-demo, echo=TRUE}
res_lm <- robustness_lm(change ~ arm + baseline_pain, data = pain_ancova_trial,
                        term = "arm", n_boot = 300, seed = 123)
res_lm$robustness_metrics[, c("worstcase_fragility_k", "worstcase_fragility_pct",
                              "bootstrap_reproducibility", "overall_robustness")]
```

The score is numeric-only: `r fmt(res_lm$robustness_metrics$overall_robustness)`
with no label, exactly as the negative result above requires. A TOST
equivalence analysis behaves the same way:

```{r tost-demo, echo=TRUE}
res_tost <- robustness_tost(pain_treatment, pain_placebo, type = "equivalence",
                            endpoint = "mean", delta_L = -20, delta_U = 20,
                            n_boot = 300, seed = 123)
c(fragility_k = res_tost$robustness_metrics$worstcase_fragility_k,
  score = round(res_tost$robustness_metrics$overall_robustness, 1))
```

A Fisher analysis inside the calibrated profile, under the calibrated weights,
does receive a label:

```{r fisher-demo, echo=TRUE}
set.seed(2026)
ctrl <- rbinom(60, 1, 0.15); trt <- rbinom(60, 1, 0.40)
res_f <- robustness_analysis(trt, ctrl, test_type = "fisher", n_boot = 1000,
                             seed = 123,
                             weights = c(jackknife = 0, fragility = 0.5,
                                         bootstrap = 0.5))
c(score = round(res_f$robustness_metrics$overall_robustness, 1),
  label = res_f$robustness_interpretation)
```

```{r fisher-default}
res_f_def <- robustness_analysis(trt, ctrl, test_type = "fisher",
                                 n_boot = 1000, seed = 123)
```

The same data under default weights returns identical component metrics but a
different composite (`r fmt(res_f_def$robustness_metrics$overall_robustness)`)
and a suppressed label --- the registry requires the calibrated weight vector,
not merely the calibrated method.

# Performance

Cost is dominated by the greedy search, which performs $O(n \cdot k_{\max})$
test evaluations --- about 3,400 tests at $n = 55$. Closed-form tests are
therefore sub-second; model engines pay for repeated refits.

```{r timingtable}
tt <- timing
tt$seconds <- round(tt$seconds, 2)
tt <- stats::reshape(tt, idvar = "engine", timevar = "n_per_arm",
                      direction = "wide")
names(tt) <- c("Engine", "n = 50 per arm (s)", "n = 150 per arm (s)")
knitr::kable(tt, row.names = FALSE, booktabs = TRUE,
  caption = "Wall-clock seconds for a single analysis with 1000 bootstrap resamples, stabilitest 0.6.0 on an Apple aarch64 laptop. The Cox benchmark uses simulated survival data generated in the timing script.")
```

The practical implication is that interactive use is comfortable at trial
scale, and that a large calibration study --- thousands of replications ---
belongs in a batch script, which is how the studies reported here were run.

# Testing and quality assurance

The package carries `r test_counts$pass` passing `testthat` assertions across
`r test_counts$files` test files, covering the analysis engines, edge cases (degenerate groups, zero
variance, complete separation, failed refits), the packaged datasets, the
calibration registry's resolution logic, and the calibration documentation
itself.

That last one is unusual enough to describe. A source-tree audit
(`tools/check-calibration-documentation.R`, run as part of the test suite)
asserts that the README, NEWS, vignettes, manuscript, and every study SAP
still state the policy invariants: that uncalibrated labels are suppressed,
that Welch 55/70 is described as a Welch comparator and not an ANCOVA
fallback, that the fail-closed reason string is the exact plural
`no_feasible_thresholds` and is not softened with "or equivalent", and that
each frozen acceptance gate still appears verbatim in its own SAP. The intent
is to make it mechanically hard to quietly weaken a published calibration
claim through documentation drift.

Each calibration study additionally publishes compact decision artifacts in
its own `published/` directory: occupancy and failure tables, power
verification, candidate diagnostics, a registry row, manifests, and a hash
ledger. Candidate hashes are recorded before held-out data is opened, so
the sequencing of any decision is auditable after the fact.

# Discussion and limitations

The framework separates three questions usually blurred together, and its
sharpest practical lesson is the gap between adversarial and symmetric
removal: a robustness claim resting on "we removed the outliers and the result
held" can overstate stability several-fold.

Several limitations deserve statement rather than burial.

**The score is largely a monotone transform of the $p$-value.** In clean
parametric settings the composite score re-expresses the evidence the
$p$-value already carries; the ANCOVA program measured this directly. This is
a scope condition, not a defect, but it bounds what the score can be used for.
Its added value lies in (i) identifying the specific observations carrying the
conclusion, (ii) the worst-case deletion bound, and (iii) behaviour under
contamination, where the $p$-value alone is silent --- see the null-with-
contamination rows of Table \@ref(tab:simtable). The v3 Track E study tested
whether the score adds detection power for *assumption violations* and found
it does not; that result is reported above rather than set aside.

**The weights are a convention.** They are explicit, pre-specifiable, and
demonstrably not load-bearing for discrimination (the Fisher weight study
found AUC essentially unchanged across weightings), but they are not derived
from theory. The score is a communication device; the components are the
substance.

**The greedy index is an upper bound.** Exact minimal overturning subsets are
combinatorially hard. Greedy search is near-optimal for monotone
single-deletion influence, and the reported trajectory lets a reader judge the
margin, but certified minimal sets --- via integer-programming formulations
--- remain future work.

**Binary significance framing.** Inherited from the fragility concept itself
and criticised for good reason [@potter2020]; the package's response is to
report the trajectory alongside the index and to frame the index as an upper
bound, wherever the index is defined rather than only in a limitations
section.

**Independence is assumed.** Clustered and longitudinal designs are not
supported; for those, `influence.ME` addresses the grouping-factor question.

**Calibration coverage is deliberately narrow.** Two configurations carry
labels. Everything else --- paired, rank, Brunner--Munzel, non-canonical
binary, model-based, Cox, TOST --- returns numeric output with a suppressed
label pending independent calibration. Users who want a verdict where none is
licensed are, by design, not given one.

# Summary

\pkg{stabilitest} asks how much of the data a conclusion actually rests on,
and answers with component metrics that name the observations responsible. Its
methodological position is that a categorical robustness verdict is a claim
requiring validation for the specific analysis at hand, and that the honest
response to a failed validation is silence plus a published negative result.
The package is available from
[github.com/ma-brain/stabilitest](https://github.com/ma-brain/stabilitest).
