---
title: "ANCOVA Case Study: Canonical Treatment Adjustment"
author: "Marius Ally"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ANCOVA Case Study: Canonical Treatment Adjustment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

```{r setup}
library(stabilitest)
```

## Purpose

This vignette reproduces the illustrative synthetic ANCOVA case study that
accompanies the independent `lm_ancova` calibration study. The dataset
`pain_ancova_trial` was prospectively frozen before production calibration and
does **not** enter training, candidate selection, or held-out validation.

## Frozen synthetic trial

`pain_ancova_trial` is an 80-row balanced two-arm trial with a continuous
baseline covariate and Week 12 pain outcome. See `?pain_ancova_trial` and
`data-raw/pain_ancova_trial.R` (seed `20260806`).

```{r data-glimpse}
str(pain_ancova_trial)
table(pain_ancova_trial$arm)
```

## Worked robustness analysis

The call below uses the exact configuration required for runtime applicability
(`n_boot = 1000`, default weights, 30% removal budget). Do not shorten the
bootstrap count merely for vignette speed.

```{r case-analysis}
case_result <- robustness_lm(
  week12_pain ~ arm + baseline_pain,
  pain_ancova_trial,
  term = "armActive",
  alpha = 0.05,
  n_boot = 1000,
  max_removal_pct = 0.30,
  weights = c(jackknife = 0.4, fragility = 0.4, bootstrap = 0.2),
  seed = 1408
)
print(case_result)
```

## How to read the output

- Numeric component metrics and the composite score are always retained.
- A categorical Fragile / Moderately Robust / Robust label appears only when
  the active `lm_ancova` registry row is validated **and** the analysis profile
  matches the supported canonical conditions.
- At Gate A the active registry remains uncalibrated, so this vignette should
  render an explicit numeric-only result. After Gate B it will report either a
  validated label or continue to suppress the label if calibration failed
  closed.

```{r profile-and-label}
case_result$analysis_profile
case_result$calibration
case_result$interpretation_label
```

## Non-calibrating status

This example is illustrative only. It must never be described as
threshold-fitting or validation evidence for the ANCOVA categorical bands.
