---
title: "Pain Trial Case Study: Interpreting Robustness Scores"
author: "Marius Ally"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pain Trial Case Study: Interpreting Robustness Scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

```{r setup}
library(stabilitest)
```

## What the package measures

`stabilitest` asks a question that a single *p*-value cannot: **how easily could this conclusion be overturned?** For a pre-specified two-sample (or model-term) analysis it combines three complementary sensitivity views:

1. **Jackknife leave-one-out** — which individual observations drive the result?
2. **Worst-case removal** — greedy adversarial deletion in the spirit of the maximum influence perturbation of Broderick, Giordano & Meager (2023). The resulting *removal fragility index* is the size of an identified subset whose removal flips the significance conclusion (an upper bound on the minimal such set).
3. **Bootstrap reproducibility probability** (Goodman, 1992) — would a replicate sample drawn from the observed effect likely reach the same conclusion?

A composite score on 0–100 summarises the components with default weights 0.4 / 0.4 / 0.2 (jackknife, worst-case fragility, bootstrap). Extreme-value (grand-mean) removal is retained as a *descriptive* companion and does **not** enter the score. Primary interpretation should stay with the component metrics; the composite is a communication heuristic calibrated by simulation (see the accompanying manuscript).

> **Status.** Experimental (v0.1.0). API and band cut-offs may evolve. Bands below apply to *significant* primary results.

## Pain trial case study

The package ships a fixed synthetic Phase II analgesic dataset used in the manuscript case study: change from baseline in pain score (0–100 scale; negative = improvement) at Week 12.

```{r data-glimpse}
length(pain_treatment)
length(pain_placebo)
summary(pain_treatment)
summary(pain_placebo)
```

Treatment includes one extreme responder (−52.0); placebo includes one minimal responder (+3.0). These are intentional for influence diagnostics, not clinical claims.

Run the Welch two-sample analysis. Bootstrap uses `n_boot = 500` here to keep vignette runtime modest; the manuscript case study used `B = 2000`. Deterministic components (jackknife, removal analyses) do not depend on `n_boot`.

```{r case-study}
res <- robustness_analysis(
  pain_treatment, pain_placebo,
  test_type = "t.test",
  n_boot = 500,
  seed = 123,
  interpret = TRUE
)
print(res)
```

Key deterministic takeaways that match the manuscript case study:

- Primary Welch test: mean difference (treatment − placebo) about −11.4, *p* ≈ 0.0018.
- Jackknife conclusion stability 100% across all leave-one-out deletions (near-perfect LOO stability mostly reflects *n* ≈ 55, not invulnerability).
- Worst-case fragility index **k = 6** (about 11% of the sample); *p* rises to ≈ 0.060 after the sixth greedy deletion.
- Overall score lands in the **Robust** band when bootstrap reproducibility is high (manuscript: ~72.5/100 with `B = 2000`).

Optional diagnostic plots (removal *p*-value trajectories and the bootstrap *p* distribution):

```{r case-plots, fig.cap = "Worst-case vs extreme-value removal trajectories (top) and bootstrap p-value distribution (bottom)."}
plot(res)
```

Set `interpret = TRUE` to attach a narrative report (`res$interpretation$report`) that restates the components and a band-based recommendation.

## Interpretation bands

With default weights, bands were calibrated by simulation *conditional on a significant result* (the setting in which the framework is usually consulted):

| Overall score | Label | Typical simulation profile |
|---------------|-------|----------------------------|
| > 70 | **Robust** | True large effects (mean score ~75) |
| 55–70 | **Moderately Robust** | True moderate effects; report components transparently |
| ≤ 55 | **Fragile** | Chance-significant findings under H₀ (mean score ~52) |

In code:

```{r bands-logic, eval = FALSE}
# As implemented in robustness_analysis():
#   > 70  -> "Robust"
#   > 55  -> "Moderately Robust"
#   else  -> "Fragile"
```

Practical reading of the components alongside the label:

- **Jackknife stability** near 100% is weak evidence of robustness in larger samples (single-observation influence shrinks as ~1/*n*). Review flagged observations when present.
- **Worst-case fragility** below ~5% of the sample should trigger clinical/data-quality review of the removal set, whatever the composite says. Median *k*_frag for chance findings in simulation was 1–2 observations versus ~5–14 for larger true effects (depending on *n* and *d*).
- **Bootstrap reproducibility** reflects strength of evidence at the *observed* effect size, not robustness to contamination: a marginal *p*-value yields low reproducibility even in clean data. That is why its default weight is only 0.2.
- **Extreme-value removal** can look much more reassuring than worst-case removal; the gap between the two indices is informative and should not be ignored.

For **non-significant** primary results the composite bands are **not** calibrated the same way — adversarial removal can manufacture significance in almost any null sample. Prefer the component metrics directly.

## Brief caveats

- The greedy fragility index is an *upper bound* on the size of a minimal overturning subset; a smaller set may exist.
- Binary significance framing inherits the usual knife-edge problem near α; always inspect the *p*-value trajectory, not only *k*_frag.
- Adversarial removal asks “could the data support the opposite conclusion?” and should complement, not replace, assumption-based sensitivity analyses (missing data, model form).
- Calibration was performed for the Welch test under normality with directional contamination; ANCOVA / Cox engines share the same scoring logic but are not yet separately calibrated.
- Pre-specify weights, `max_removal_pct`, and bands in the SAP when used in confirmatory work.

Further methodological detail, simulation tables, and reporting templates are in the accompanying manuscript / methodological notes (development source tree and related project materials). For model-based terms see `robustness_lm()` and `robustness_surv()`.
