---
title: "Pain Trial Case Study: Interpreting Robustness Scores"
author: "Marius Ally"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pain Trial Case Study: Interpreting Robustness Scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

```{r setup}
library(stabilitest)
```

## What the package measures

`stabilitest` asks a question that a single *p*-value cannot: **how easily could this conclusion be overturned?** For a pre-specified two-sample (or model-term) analysis it combines three complementary sensitivity views:

1. **Jackknife leave-one-out** — which individual observations drive the result?
2. **Worst-case removal** — greedy adversarial deletion in the spirit of the maximum influence perturbation of Broderick, Giordano & Meager (2023). The resulting *removal fragility index* is the size of an identified subset whose removal flips the significance conclusion (an upper bound on the minimal such set).
3. **Bootstrap reproducibility probability** (Goodman, 1992) — would a replicate sample drawn from the observed effect likely reach the same conclusion?

A composite score on 0–100 summarises the components with default weights 0.4 / 0.4 / 0.2 (jackknife, worst-case fragility, bootstrap). Extreme-value (grand-mean) removal is retained as a *descriptive* companion and does **not** enter the score. Numeric scores and component metrics are retained for every supported method. Categorical labels are suppressed for uncalibrated methods and conclusions; only an applicable significant `welch_unpaired` result receives the Robust / Moderately Robust / Fragile label. The public `robustness_analysis()` dispatcher is unchanged. `lm_ancova` is the next independent calibration target.

> **Status.** Experimental (v0.5.1). API and calibration metadata may evolve. The current 55/70 bands apply only to the documented significant Welch configuration. The broader Task 15 simulation is historical evidence, not an active calibration for other methods.

## Pain trial case study

The package ships a fixed synthetic Phase II analgesic dataset used in the manuscript case study: change from baseline in pain score (0–100 scale; negative = improvement) at Week 12.

```{r data-glimpse}
length(pain_treatment)
length(pain_placebo)
summary(pain_treatment)
summary(pain_placebo)
```

Treatment includes one extreme responder (−52.0); placebo includes one minimal responder (+3.0). These are intentional for influence diagnostics, not clinical claims.

Run the Welch two-sample analysis. Bootstrap uses `n_boot = 500` here to keep vignette runtime modest; the manuscript case study used `B = 2000`. Deterministic components (jackknife, removal analyses) do not depend on `n_boot`.

```{r case-study}
res <- robustness_analysis(
  pain_treatment, pain_placebo,
  test_type = "t.test",
  n_boot = 500,
  seed = 123,
  interpret = TRUE
)
print(res)
```

Key deterministic takeaways that match the manuscript case study:

- Primary Welch test: mean difference (treatment − placebo) about −11.4, *p* ≈ 0.0018.
- Jackknife conclusion stability 100% across all leave-one-out deletions (near-perfect LOO stability mostly reflects *n* ≈ 55, not invulnerability).
- Worst-case fragility index **k = 6** (about 11% of the sample); *p* rises to ≈ 0.060 after the sixth greedy deletion.
- Overall score lands in the **Robust** band when bootstrap reproducibility is high (historical manuscript reference: ~72.5/100 with `B = 2000`; the label is applicable here because this is the documented Welch configuration).

Optional diagnostic plots (removal *p*-value trajectories and the bootstrap *p* distribution):

```{r case-plots, fig.cap = "Worst-case vs extreme-value removal trajectories (top) and bootstrap p-value distribution (bottom)."}
plot(res)
```

Set `interpret = TRUE` to attach a narrative report (`res$interpretation$report`) that restates the components and, when calibration is applicable, a band-based recommendation. For an uncalibrated result, the report retains numeric information but suppresses the categorical label.

## Choosing an analysis

All public entry points apply the same three-part robustness framework, but the
primary test and the unit removed or resampled differ. Choose the function that
matches the pre-specified analysis rather than selecting one from the resulting
robustness score.

| Analysis question | Entry point | Primary test represented in the robustness calculation |
|-------------------|-------------|---------------------------------------------------------|
| Independent or paired continuous groups | `robustness_analysis()` | Welch or paired *t* test |
| Rank-based comparison | `robustness_analysis()` | Wilcoxon rank-sum/signed-rank, or unpaired Brunner–Munzel |
| Two independent binary groups | `robustness_analysis()` | Fisher exact, chi-square, or `prop.test` |
| Linear model / ANCOVA term | `robustness_lm()` | Coefficient *t* test or multi-df joint F test |
| Binary or count regression term | `robustness_glm()` | Binomial-logit or Poisson-log coefficient test / joint LRT |
| Cox proportional-hazards term | `robustness_surv()` | Coefficient Wald test or multi-df joint LRT |
| Equivalence or non-inferiority | `robustness_tost()` | Mean difference, risk difference, or odds-ratio TOST/NI |

Representative calls are shown below. They are not evaluated in this case-study
vignette so that it remains quick to build.

```{r analysis-selection, eval = FALSE}
# Rank tests: Brunner–Munzel is unpaired and allows unequal variances.
rank_res <- robustness_analysis(
  group1, group2, test_type = "brunner_munzel", n_boot = 1000
)

# Individual-level binary outcomes, coded 0/1 or logical.
binary_res <- robustness_analysis(
  responders_active, responders_control,
  test_type = "fisher", n_boot = 1000
)

# A coefficient name requests its usual single-df test.
ancova_res <- robustness_lm(
  change ~ arm + baseline, trial_data,
  term = "armActive", n_boot = 1000
)

# A multi-level term label requests the corresponding joint test.
ancova_joint <- robustness_lm(
  change ~ arm + baseline, trial_data,
  term = "arm", n_boot = 1000
)

# Supported GLMs use family objects. Composite score weights and observation
# weights are separate arguments: weights versus obs_weights.
glm_res <- robustness_glm(
  response ~ arm + baseline, trial_data,
  term = "armActive", family = binomial(),
  obs_weights = trial_data$analysis_weight, n_boot = 1000
)
poisson_res <- robustness_glm(
  events ~ arm + offset(log(exposure)), event_data,
  term = "arm", family = poisson(), n_boot = 1000
)

cox_res <- robustness_surv(
  survival::Surv(time, event) ~ arm + age, survival_data,
  term = "armActive", n_boot = 1000
)

# Equivalence uses both one-sided tests; NI uses the relevant one-sided test.
equivalence_res <- robustness_tost(
  group1, group2, type = "equivalence", endpoint = "mean",
  margin = 0.5, n_boot = 1000
)
ni_or_res <- robustness_tost(
  responders_active, responders_control,
  type = "noninferiority", endpoint = "or", margin = 1.25,
  higher_is_better = TRUE, n_boot = 1000
)
```

For rank tests, the reported location effect is the Hodges–Lehmann shift;
Brunner–Munzel also records stochastic superiority. Binary TOST/NI currently
uses Wald risk-difference or log-odds-ratio methods, which can be
anticonservative with sparse tables. See each function's help page for its
assumptions and endpoint-specific limitations.

## Reliability checks and result diagnostics

Current releases fail explicitly when a robustness component cannot support a
meaningful calculation:

- A removal analysis must have capacity for at least one deletion while
  retaining the engine's minimum sample size. A no-flip value of `max_k + 1`
  is a right-censored result and is reported only after the full configured
  deletion horizon has been evaluated; it does not mean the conclusion can
  never be overturned.
- Two-sample bootstrap attempts with errors or non-finite *p*-values remain in
  `res$bootstrap$results`. Finite replicates determine the summaries, while
  `res$bootstrap$n_valid` and `res$bootstrap$n_failed` make the accounting
  visible. An entirely degenerate bootstrap raises an error.
- `robustness_lm()`, `robustness_glm()`, and `robustness_surv()` apply every
  deletion and resample to exactly the rows retained by the full fitted model.
  Their reported `n` therefore excludes rows omitted for missing model
  variables; GLM `obs_weights` stay aligned with the retained original rows.
- Composite `weights` must be a named numeric vector containing exactly
  `jackknife`, `fragility`, and `bootstrap`, with finite non-negative values
  that sum to one. These are distinct from GLM observation weights.

The two-sample and model-based result classes expose aligned aliases for common
fields: `metrics` / `robustness_metrics`, `interpretation_label` /
`robustness_interpretation`, and `original_estimate` / `original_mean_diff`.
Component objects remain class-specific, so inspect `str(result)` or the
function's `Value` help section before writing downstream extraction code.

## Interpretation policy

The numeric score and all component metrics remain available regardless of the
test family, endpoint, or conclusion. The active runtime registry has one
validated row: significant `welch_unpaired` with the default score definition,
weights, and documented independent-groups conditions. Only that row uses the
55/70 labels:

| Overall score | Label | Typical simulation profile |
|---------------|-------|----------------------------|
| > 70 | **Robust** | Narrow Welch calibration |
| (55, 70] | **Moderately Robust** | Narrow Welch calibration |
| ≤ 55 | **Fragile** | Narrow Welch calibration |

Exact 70 is Moderately Robust; exact 55 is Fragile.

In code:

```{r bands-logic, eval = FALSE}
# As implemented for an applicable significant welch_unpaired result:
#   > 70  -> "Robust"              # score > 70
#   > 55  -> "Moderately Robust"   # (55, 70]
#   else  -> "Fragile"             # score ≤ 55
```

Practical reading of the components alongside the label:

- **Jackknife stability** near 100% is weak evidence of robustness in larger samples (single-observation influence shrinks as ~1/*n*). Review flagged observations when present.
- **Worst-case fragility** below ~5% of the sample should trigger clinical/data-quality review of the removal set, whatever the composite says. Median *k*_frag for chance findings in simulation was 1–2 observations versus ~5–14 for larger true effects (depending on *n* and *d*).
- **Bootstrap reproducibility** reflects strength of evidence at the *observed* effect size, not robustness to contamination: a marginal *p*-value yields low reproducibility even in clean data. That is why its default weight is only 0.2.
- **Extreme-value removal** can look much more reassuring than worst-case removal; the gap between the two indices is informative and should not be ignored.

For every other method-specific unit (including paired t, rank tests, binary
tests, model terms, Cox, and TOST) and for non-significant or otherwise
inapplicable conclusions, `interpretation_label` is `NA`. Prefer the numeric
score and component metrics directly. Task 15's broad-family 55/70 tables are
archived under `manuscript/calibration/published/` as historical evidence only.

## Brief caveats

- The greedy fragility index is an *upper bound* on the size of a minimal overturning subset; a smaller set may exist.
- Binary significance framing inherits the usual knife-edge problem near α; always inspect the *p*-value trajectory, not only *k*_frag.
- Adversarial removal asks “could the data support the opposite conclusion?” and should complement, not replace, assumption-based sensitivity analyses (missing data, model form).
- Calibration is retained only for the narrowly documented Welch test under normality with directional contamination. ANCOVA, GLM, Cox, proportion, rank, and TOST/NI engines share the scoring logic but remain uncalibrated; `lm_ancova` is the next independent calibration project.
- Pre-specify weights, `max_removal_pct`, and bands in the SAP when used in confirmatory work.

Further methodological detail, simulation tables, and reporting templates are in the accompanying manuscript / methodological notes (development source tree and related project materials). For model-based terms see `robustness_lm()`, `robustness_glm()`, and `robustness_surv()`.
