---
title: "Proportion Test Case Study: Binary Response in an Oncology Trial"
author: "Marius Ally"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Proportion Test Case Study: Binary Response in an Oncology Trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

```{r setup}
library(stabilitest)
```

## Scope of this vignette

This vignette illustrates the two-group binary-proportion workflow
(`robustness_analysis(test_type = "fisher")`) and what the published
`fisher_exact` calibration does and does not license. The calibration applies
**only** to eligible significant canonical two-arm Fisher exact results; every
other proportion result retains numeric scores and component metrics while the
categorical label stays suppressed.

The calibration was prospectively registered before any held-out data were
opened. The single cutoff `L = 58` was fit on training and confirmed on a
fresh held-out grid (conservative of Wilson and scenario-cluster bootstrap
bounds). Details live in the study tree at
`manuscript/calibration/studies/binary_proportion/`.

## The frozen case-study dataset

`onc_response_trial` is a prospectively frozen synthetic two-arm oncology
responder trial (60 patients per arm, binary objective response). It was
generated once from a committed generator with seed `20260809` before any
p-value, score, or band was inspected, and it is **not** part of the
calibration training or held-out evidence. We use it here exactly once, after
the calibration was published, to illustrate the honest label state.

```{r data}
active <- onc_response_trial$response[onc_response_trial$arm == "Active"]
control <- onc_response_trial$response[onc_response_trial$arm == "Placebo"]

# Group sizes and observed response rates
c(active = sum(active), control = sum(control))
c(active_rate = mean(active), control_rate = mean(control))
```

## Primary analysis

We run the primary Fisher exact test. Following the frozen calibration design,
the robustness analysis uses the jackknife-light weights
(`fragility = 0.5`, `bootstrap = 0.5`, `jackknife = 0`), `n_boot = 1000`, and a
documented seed.

```{r analysis}
set.seed(20260811)
res <- robustness_analysis(
  active, control,
  test_type = "fisher",
  weights = c(jackknife = 0, fragility = 0.5, bootstrap = 0.5),
  n_boot = 1000, seed = 20260811
)

res$original_p
res$original_significant
res$original_mean_diff
```

The primary Fisher test is **not significant** (`p = `r signif(res$original_p, 3)`).
The observed active minus control response-rate difference is
`r round(res$original_mean_diff, 2)`.

## Three components

Even without a significant conclusion, `stabilitest` reports the three
component views that summarise how easily the result moves:

```{r components}
m <- res$robustness_metrics
c(
  jackknife_stability = m$jackknife_conclusion_stability,
  worstcase_fragility_component = m$worstcase_fragility_component,
  worstcase_fragility_k = m$worstcase_fragility_k,
  bootstrap_reproducibility = m$bootstrap_reproducibility,
  overall_score = m$overall_robustness
)
```

- **Jackknife leave-one-out** (`jackknife_stability`) records conclusion
  stability under single-observation deletion.
- **Worst-case removal fragility** (`worstcase_fragility_k`) is the size of the
  greedy adversarial subset whose deletion moves the p-value furthest toward
  overturning the conclusion; the rescaled component is
  `worstcase_fragility_component`.
- **Bootstrap reproducibility** (`bootstrap_reproducibility`) estimates the
  probability a replicate sample reaches the same conclusion at the observed
  effect.

## Walsh event-flip fragility vs removal fragility

A classic comparator for binary outcomes is the **Walsh event-flip fragility
index**: the number of outcome flips (0 to 1 in the smaller-event arm) needed
to change the Fisher conclusion. It is distinct from the removal-based
fragility component that enters the score:

```{r walsh}
# Walsh FI: 0 here because the result is already non-significant (nothing to
# overturn). For a significant result it counts flips toward p >= alpha.
walsh_fi <- 0L
c(walsh_event_flip_fi = walsh_fi, removal_fragility_k = m$worstcase_fragility_k)
```

For this non-significant result the Walsh index is `0` (there is no
significance to overturn), while the removal-based index records how many
greedy deletions move the p-value. The two answer different questions and are
kept separate: the Walsh index is a literature comparator archived per
replicate; only the removal-based component enters the composite score.

## What the published calibration licenses

The published `fisher_exact` calibration emits a categorical band **only** for
an eligible significant canonical two-arm result. Because this case study is
non-significant, no band applies:

```{r label}
res$calibration$status
res$calibration$applicable
res$interpretation_label
```

This is the intended fail-closed behaviour: a non-significant conclusion is
`bands_not_applicable`, the label stays `NA`, and numeric scores and component
metrics remain fully available. Categorical labels are never manufactured for
conclusions the calibration was not built to interpret.

> **Status.** The active runtime registry still lists `fisher_exact` as
> `uncalibrated`; the published `validated_method_specific` decision and its
> cutoff live in the study publication artefacts and are activated in the
> runtime registry only under separate human-approved work (Gate B).

## Eligibility recap

A `fisher_exact` result receives a band only when all of the following hold:
two-arm individual-level binary input, complete cases, a significant
conclusion, per-arm *n* within [25, 200], allocation ratio within [0.8, 1.25],
an observed control-arm event rate within [0.08, 0.55] with at least 3 events
and 3 non-events, and the frozen analysis configuration
(`alpha = 0.05`, `n_boot = 1000`, `max_removal_pct = 0.30`, jackknife-light
weights). Everything else is numeric-only.
